---
title: Functional Programming with `R`
author: "Miles D. Williams"
date: "2019 UIUC Political Science Math Camp"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Now that you've gained some familiarity with loops, let's add just another layer of complexity. This won't hurt too badly, I promise...

`R` is a functional programming language. In short, that means that you can create your own functions in `R` that will do basically whatever you tell them to do. (Word to the wise, it will also do whatever stupid thing you tell it to do)

This might seem like an odd feature. Why write a new function? `R` already comes with so many nice built-in functions; adding our own may appear silly.

But, consider for a moment the value of the `for` loop. It allows you to automate an iterative task. It takes what would otherwise require 1,000 lines of code and condenses it to a mere handful. Functional programming allows you to do much the same thing. In fact, you can write functions that perform loops for you, without having to rewrite a loop each time you want to do an iterative task.

Consider the following simple loop:

```{r}

# We first create a 10 x 10 matrix of zeros
x = matrix(0, ncol = 10, nrow = 10)
x # Show matrix

# We then fill each column of this matrix with random variables, but
# with the mean centered at a value equal to the column number,
# e.g., for column 1, the mean of the random variable to be generated will be 1.

# The loop:
for(i in 1:ncol(x)){
  x[,i] = rnorm(n = nrow(x), sd = 1, mean = i)
}
```

Here's the output from the loop:

```{r}
x
```

Now let's say we want to be able to specify what the mean will be in each column, and for that matter the standard deviation:

```{r}
x = matrix(0, ncol = 10, nrow = 10)
means = c(4,9,3,4,0,-1,8,6,2,5)
sds = c(.3,8,9,.5,2,1,6,3,7,2)
for(i in 1:ncol(x)){
  x[,i] = rnorm(n = nrow(x), sd = sds[i], means[i])
}
x
```

Now, don't get me wrong, I love `R`, but writing all that out was tedious. I don't really want to rewrite all that every time I want to generate a matrix of random variables...functional programming to the rescue!

`R` lets you write over a particular routine and save that routine as a function. This is done straightforwardly with the `function()` command. For example:

```{r}
# My mean function:
my_mean = function(x){
  x_bar = sum(x)/length(x)
  return(x_bar)
}

# Let's take the mean of the first element of the x matrix
my_mean(x[,1])

# How does this compare with R's built-in mean funciton?
mean(x[,1])
```

The basic outline of writing a function is simple:
  
  1. specify the name of the function you want to create:

  - `my_mean =` (this can also be done with the longer `<-`)
  
  2. Use `function()` and place inside the object you want the function to work on:
  
  - `my_mean = function(x)` (x is just a stand-in for the variable we want to take the mean of)
  
  3. Then, in brackets, we specify the routine we want the function to automate:
  
  - `my_mean = function(x) { ... }`
  
  
Pretty simple right? Can we apply this to the loop example above? Let's give it a go.

```{r}
random_mat = function(
  n, # number of rows
  k  # number of columns
) {
  
  # Make empty n x k matrix
  x = matrix(0, ncol = k, nrow = n)
  
  # Have routine pick a random mean per column
  means = rnorm(n = k)
  
  # Have routine pick a random sd per column
  sds = exp(rnorm(n = k))
  
  # Fill the matrix
  for(i in 1:k){
    x[,i] = rnorm(n = n, sd = sds[i], means[i])
  }
  
  # Have the function return the filled matrix
  return(x)
}
```

Does it work? Set `n = 10` and `k = 3`:

```{r}
random_mat(n = 10, k = 3)
```

It works perfectly. 

# Practice
Now it's your turn. 

  1. Complete the following routine with the correct line of code. Simply delete the `#` in front of the lines of code you think should be in the function. Explain.
  
```{r eval=F}
# This function will plot multiple histograms for a matrix
# of covariates. Which lines of code will correctly plot them?
multi_hist = function(x) {
  
  #par(mfcol = c(1,nrow(x))) # Are these two lines correct?
  #for(i in 1:nrow(x)){  
    
  #par(mfcol = c(1,ncol(x))) # Or are these two lines correct?
  #for(i in 1:ncol(x)){
    hist(x[,i],main=paste("Variable", i),xlab="Values")
  }
}

# Test it
multi_hist(x=random_mat(n = 100, k = 3))
```

  2. In applied settings, it is common to "standardize" variables. That is, to center covariates at 0 and scale them to what are called *standard deviation* units. The math for doing this is fairly simple. For a variable $x_n$: $\frac{x_n - \bar{x}}{\text{sd}(x_n)}$. In other words, for $n$ values of a random variable $x$, from each value subtract the mean of the variable then divide by the standard deviation of the variable. Finish the standardization function below:
  
```{r eval = F}
stand = function(x){
  # What goes here?
}
```

  3. Using the `stand()` function you completed above, can you write a function around the following loop that standardizes each variable in a matrix?
  
```{r eval = F}

# Make empty matrix
stand_mat = matrix(0, ncol = ncol(x), nrow = nrow(x))

# Use loop to fill matrix with standardized values
for(k in 1:ncol(x)){
  stand_mat[,k] = stand(x[,k])
}

```

  
  
  